UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-08-16 ms/main.tex
manuscript changes, add some R scripts used in analysis and dataviz

\begin{abstract}
{\large Abstract.}
%The abstract is typically 100-150 words, unreferenced. 
Human populations feature both discrete and continuous patterns of variation. Current analysis approaches struggle to jointly identify these patterns because of modelling assumptions, mathematical constraints, or numerical challenges. Here we apply uniform manifold approximation and projection (UMAP), a non-linear dimension reduction tool, to three well-studied genotype datasets and discover overlooked subpopulations within the American Hispanic population, fine-scale relationships between geography, genotypes, and phenotypes in the UK population, and cryptic structure in the Thousand Genomes Project data. This approach outperforms other visualization methods, is well-suited to the influx of large and diverse data, and opens new lines of inquiry in population-scale datasets.
\end{abstract}

@@

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
\subsection*{Author summary}
How about: The demographic history of human populations features evolving geographic and social barriers to mating. Over time, these barriers have led to varying levels of genetic relatedness among individuals.  This \emph{population structure} is informative about human history, and can have a significant impact on studies of medical genetics. Because population structure depends on myriad demographic forces, \emph{a priori} visualization is useful to identify subtle patterns of population structure. We use a dimension reduction method --- UMAP --- to visualize population structure in three genomic datasets and find previously unobserved patterns, revealing fine-scale population structure and illustrating differences between groups in traits such as white blood cell count, height, and FEV1, a measure used to detect airway obstruction. Using UMAP is computationally efficient and can identify fine-scale population structure in large population datasets. We find it particularly useful to reveal phenotypic variation among genetically related populations, and recommend it is a complement  to principal component analysis in primary data visualization.

@@

Here we will focus on nonparametric approaches to visualize relatedness patterns among individuals within populations. If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of copies of the derived alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ ranging from hundreds of thousands to hundreds of millions. Since each individual is represented as an $L$-dimensional vector, dimension reduction methods are needed to visualize the data.

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=0.7\textwidth]{images/montage_1000G_labels_axes.pdf}
    \caption{\textbf{Four methods of dimension reduction of 1KGP genotype data with population labels.} (i) PCA maps individuals in a triangle with vertices corresponding to African, Asian, and European continental ancestry. Discarding lower-variance PCs leads to overlap of populations with no close affinity, such as Central and South American populations with South Asians. (ii) t-SNE forms groups corresponding to continents, with some overlap between European and Central and South American people. Smaller subgroups are visible within continental clusters. The cloud of peripheral points results from the method's poor convergence. (iii) UMAP forms distinct clusters related to continent with clearly defined subgroups. Japanese, Finnish, Luhya, and some Punjabi and Telugu populations form separate clusters consistent with their population history\cite{10002015global}. (iv) UMAP on the first 15 principal components forms fine-scale clusters for individual populations. Groups closely related by ancestry or geography, such as African Caribbean/African American, Spanish/Italian, and Kinh/Dai populations cluster together. Results using t-SNE on principal components are presented in figure~\ref{fig:supp_megamontage_pc2_9}. Axes in UMAP and t-SNE are arbitrary. Since the algorithms prioritize local distances, long distances between clusters are not meaningful.
    ACB, African Caribbean in Barbados;
    ASW, African Ancestry in Southwest US;
    BEB, Bengali;
    CDX, Chinese Dai;
    CEU, Utah residents with Northern/Western European ancestry;
    CHB, Han Chinese;
    CHS, Southern Han Chinese;
    CLM, Colombian in Medellin, Colombia;
    ESN, Esan in Nigeria;
    FIN, Finnish;
    GBR, British in England and Scotland;
    GWD, Gambian;
    GTH, Gujarati;
    IBS, Iberian in Spain;
    ITU, Indian Telugu in the UK;
    JPT, Japanese;
    KHV, Kinh in Vietnam;
    LWK, Luhya in Kenya;
    MSL, Mende in Sierra Leone;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PJL, Punjabi in Lahore, Pakistan;
    PUR, Puerto Rican;
    STU, Sri Lankan Tamil in the UK;
    TSI, Tuscani in Italy;
    YRI, Yoruba in Nigeria}
    \label{fig:1000g_compare}
\end{figure*}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{0.49\columnwidth}
    \includegraphics[width=\columnwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.jpeg}
    \caption{}
    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
    \end{subfigure}\hfill
    \unskip\ \vrule\
        \begin{subfigure}{0.49\columnwidth}
    \includegraphics[width=\columnwidth]{images/Asthma_FullAsian_PC_QC_new_UMAP_PC5_NC2_NN15_MD05_201931173256_ethcob.jpeg}
    \caption{}
        \label{fig:umap_ukbb_indian}
    \end{subfigure}
    \caption{\textbf{Applying UMAP to subsets of data can reveal deep population structure.} \textbf{(a)} UMAP on the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster. Colouring the points by birthplace shows they were born almost entirely in the Mountain region (in green) of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana). When populations from the 1KGP are projected onto the UMAP embedding they do not map to the cluster. Six 1KGP populations are presented:
    CLM, Colombian in Medellin, Colombia;
    IBS, Iberian in Spain;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PUR, Puerto Rican;
    TSI, Tuscani in Italy. Figures~\ref{fig:supp_umap_hrs_hisp_admix} and \ref{fig:supp_umap_hrs_hisp_birth} present the same projection of individuals from the HRS coloured by estimated admixture proportions census region of birth, respectively. \textbf{(b)} UMAP on the top $8$ principal components of the self-identified Asian populations of the UKBB create clusters. Indian individuals born in Kenya (in purple) form one such cluster. A version coloured by self-identified ethnicity is presented in figure~\ref{fig:supp_umap_ukbb_asian_eth}. }
    \label{fig:subpops}
\end{figure*}

\subsection*{Results}
\subsubsection*{Fine-scale visualization of the 1KGP dataset} The 1KGP contains genotype data of 3,450 individuals from 26 relatively distinct labeled populations\cite{10002015global}. Figure~\ref{fig:1000g_compare} shows visualizations using PCA, t-SNE, UMAP, and UMAP with PCA pre-processing. Using UMAP and t-SNE on the genotype data presents clusters that are roughly grouped by continent, with UMAP showing a clear hierarchy of population and continental clusters, whereas t-SNE fails to assign many individuals to population clusters. Using either method on the top principal components leads to distinct population clusters and less defined continental structure. Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; adding further components converges to results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}). To investigate the population information contained in low-variance PCs, we performed UMAP on data projected onto PCs 100 to 3450 (i.e., without information about the leading 99 PCs). Figure~\ref{fig:supp_1kgp_3350} shows that population structure is still clearly visible.

Focusing on UMAP with the leading 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), several population clusters reflect shared ancestries. British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry. Toscani and Iberian individuals form a cluster reflecting their Mediterranean heritage. African Americans in the Southwest US, African Caribbean individuals in Barbados, and some Puerto Ricans also form a cluster. Three East Asian clusters appear: one is largely Han and Southern Han individuals, another is comprised of the Chinese Dai in southern China and the Kinh from Vietnam, and the third is the Japanese population. Other clusters are comprised of Colombians and Peruvians, the Esan and Yoruba populations of Nigeria, and several South Asian populations.

Within population clusters, family members were projected near each other within broader population groups. When UMAP was parameterized to use only 5 nearest neighbours, however, families often formed distinct clusters (figure~\ref{fig:supp_1kgp_families}): Using few neighbors to build a manifold emphasizes closer relatedness.

A few individuals cluster with populations different from their label: some Mexican individuals cluster with Spanish and Italian populations; some Puerto Rican individuals cluster with African American and Caribbean populations; and one Gambian individual clusters with the Mende of Sierra Leone. Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering effect is robust to the number of components considered (figure~\ref{fig:supp_megamontage_pc10_50}). Differentiation in the 1KGP Gujarati population has been previously identified through a PCA restricted to the Gujarati \cite{reich2009india}. Following a preprint version of the present article, 23andme released a statement \cite{23andme} arguing that one of the two clusters could be traced, via 23andMe participant recontact to individuals from a group in Western India with shared ancestry and patronym. [D. Poznik, 23andme, personal communication, and \cite{23andme}].

@@

The HRS forms several large clusters, reflecting both ethnicity (figure~\ref{fig:supp_umap_hrs_eth}) and admixture proportions (figure~\ref{fig:umap_hrs_admix}). Gradients in admixture proportion are visible within the predominantly Hispanic cluster, but not within the predominantly Black cluster, perhaps because the variance in ancestry proportions is greater among Hispanics. The "White Not Hispanic" (WNH) group forms several interconnected clusters, and these do not correspond to broad geographical areas (figure~\ref{fig:supp_hrs_born}). By generating the PC axes and UMAP embedding for the HRS data in figure~\ref{fig:supp_umap_hrs_eth}, and projecting the 1KGP data onto it, we reveal substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population (figure~\ref{fig:supp_hrs_1kgp_projected}). One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.

\subsubsection*{Regional patterns in the Hispanic subpopulation} Applying UMAP to self-identified Hispanic individuals in the HRS reveals clear groupings related to birth region. One cluster, highlighted in figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States. This cluster is not apparent when looking at a grid of pairwise plots of the first 8 principal components, provided in figure~\ref{fig:supp_hrs_hisp_grid}, as the signal is distributed along PCs 3, 4, and 6. Even though continental admixture patterns do correlate with UMAP position (figure~\ref{fig:supp_umap_hrs_hisp_admix}), these do not explain the Mountain Region cluster. Individuals from 1KGP populations do not appear in the cluster when projected to the UMAP embedding. The cluster possibly comprises the Hispano/Nuevomexicano population of the Southwest US, who have been present in the Mountain Region area long before the more recent immigrants from Latin America, and whose ancestry is expected to reflect both distinct Native ancestry and population-specific drift relative to other Hispanic populations. Such a cluster has been previously identified in AncestryDNA data using sequential clustering on identity-by-descent connections\cite{han2017clustering}; a recent preprint discusses the Mountain Region Hispanics with a more detailed historical description\cite{Jordan333609}.

\subsubsection*{Population structure in the UKBB reflects local and global genetic variation} 
The UKBB contains data on 488,377 individuals including genotypes, phenotypic measures and self-identified ethnic backgrounds. Figure~\ref{fig:fig_ukbb} compares UMAP to PCA applied to UKBB. As expected, PCA captures major axes of variation emphasizing continental ancestry, whereas UMAP reveals finer sturcture. UMAP on the top 10 principal components reveals continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies multiple sub-populations, as well as continuous structure within populations and admixture gradients between populations. The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of a large and multi-ethnic dataset.

@@

The largest cluster in figure~\ref{fig:umap_ukbb_pc10} consists of the White British and Irish populations. The Irish population forms a sub-cluster, but many individuals are also scattered throughout the British-identifying population. Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters. Chinese individuals form a cluster, within a broader East Asian population; Indian, Pakistani, and Bangladeshi populations form a closely bound cluster as well. The East Asian and South Asian populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group". The patchwork of genetic neighbourhoods is connected by trails of admixed individuals, which converge at a nexus of individuals with a variety of ethnicities. Many claim mixed ancestry, and there are clusters of individuals who belong to an "other ethnic group". Using data on countries of birth, we identified many finer groups in figure~\ref{fig:supp_ukbb_cob}, and confirmed they appeared in intuitive areas with, e.g., Japanese and Filipino clusters being projected near Chinese clusters.

Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection from figure~\ref{fig:umap_ukbb_pc10} coloured instead by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly. Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe and British-wide data\cite{novembre2008europe,leslie2015fine}. Figure~\ref{fig:umap_ukbb_pc10} shows that the admixed individuals have UMAP coordinates next to White British individuals residing in the South East of the UK, where London is located (see also figure~\ref{fig:supp_london_distance}, where individuals are coloured by distance from London). This likely reflects the high migration levels to the city and surrounding area: the UMAP projection attempted to preserve both the genetic similarity among admixed individuals and the relatedness with White British individuals in cosmopolitan areas.

 The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved. Figures~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB.

\begin{figure}
    \centering
    \begin{subfigure}{\columnwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ns_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ew_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \caption{}
    \label{fig:umap_ukbb_geo_osgb}
    \end{subfigure}
     \begin{subfigure}{\columnwidth}
         \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_f.pdf}
    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
   \caption{}
    \label{fig:umap_height_female}
      \end{subfigure}
      \caption{\textbf{UMAP captures relationships between population structure, geography, and phenotype heterogeneity.} \textbf{(a)} Each individual is coloured by their geographical coordinates of residence. Coordinates follow the UKBB's OSGB1936 geographic grid system and represent distance from the Isles of Scilly, which lie southwest of Great Britain. The left image colours individuals by their north-south ("northing") coordinates, and the right image colours them by their east-west ("easting") coordinates. Adding more components creates finer clusters.  (figures~\ref{fig:supp_montage_ukbb_ns} and~\ref{fig:supp_montage_ukbb_ew}). Northing values were truncated between $100$km and $700$km, and easting values were truncated between $200$km and $600$km. \textbf{(b)} Females from the same projection coloured by age-adjusted difference from mean population height (left) and leukocyte counts (right). Individuals with missing data were excluded. To protect participant privacy, data in these images has been randomized as explained in the materials and methods section.}
\end{figure}

As an alternate visualization of geography and genetic diversity, we performed a 3D UMAP projection and converted the UMAP coordinates into RGB values, allowing us to plot individuals on a map of Great Britain, emphasizing both spatial gradients of genetic relatedness and increased diversity in urban centers (figure~\ref{fig:umap_ukbb_geo_map}). The geographical patterns outside major urban centers are similar to those reported in~\cite{leslie2015fine} using the haplotype-based CHROMOPAINTER on British individuals whose grandparents lived nearby. Using data about country of birth, we performed a similar analysis of a world map~\ref{fig:umap_world_map}, revealing subtle regional variation around the world.

>>>>

\begin{abstract}
{\large Abstract.}
%The abstract is typically 100-150 words, unreferenced. 
Human populations feature both discrete and continuous patterns of variation. Current analysis approaches struggle to jointly identify these patterns because of modelling assumptions, mathematical constraints, or numerical challenges. Here we apply uniform manifold approximation and projection (UMAP), a non-linear dimension reduction tool, to three well-studied genotype datasets and discover overlooked subpopulations within the American Hispanic population, fine-scale relationships between geography, genotypes, and phenotypes in the UK population, and cryptic structure in the Thousand Genomes Project data. This approach is well-suited to the influx of large and diverse data and opens new lines of inquiry in population-scale datasets.
%This approach outperforms other visualization methods, is well-suited to the influx of large and diverse data, and opens new lines of inquiry in population-scale datasets.
\end{abstract}

@@

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
\subsection*{Author summary}
The demographic history of human populations features varying geographic and social barriers to mating. Over time, these barriers have led to varying levels of genetic relatedness among individuals.  This \emph{population structure} is informative about human history, and can have a significant impact on studies of medical genetics. Because population structure depends on myriad demographic and ecological forces, \emph{a priori} visualization is useful to identify subtle patterns of population structure. We use a dimension reduction method --- UMAP --- to visualize population structure in three genomic datasets and find previously unobserved patterns, revealing fine-scale population structure and illustrating differences between groups in traits such as white blood cell count, height, and FEV1, a measure used to detect airway obstruction. Using UMAP is computationally efficient and can identify fine-scale population structure in large population datasets. We find it particularly useful to reveal phenotypic variation among genetically related populations, and recommend it is a complement  to principal component analysis in primary data visualization.

@@

Here we will focus on nonparametric approaches to visualize relatedness patterns among individuals within populations. If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of copies of the alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ ranging from hundreds of thousands to hundreds of millions. Since each individual is represented as an $L$-dimensional vector, dimension reduction methods are needed to visualize the data.

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=0.7\textwidth]{images/montage_1000G_labels_axes.pdf}
    \caption{\textbf{Four methods of dimension reduction of 1KGP genotype data with population labels.} (i) PCA maps individuals in a triangle with vertices corresponding to African, Asian, and European continental ancestry. Discarding lower-variance PCs leads to overlap of populations with no close affinity, such as Central and South American populations with South Asians. (ii) t-SNE forms groups corresponding to continents, with some overlap between European and Central and South American people. Smaller subgroups are visible within continental clusters. The cloud of peripheral points results from the method's poor convergence. (iii) UMAP forms distinct clusters related to continent with clearly defined subgroups. Japanese, Finnish, Luhya, and some Punjabi and Telugu populations form separate clusters consistent with their population history\cite{10002015global}. (iv) UMAP on the first 15 principal components forms fine-scale clusters for individual populations. Groups closely related by ancestry or geography, such as African Caribbean/African American, Spanish/Italian, and Kinh/Dai populations cluster together. Results using t-SNE on principal components are presented in fig~\ref{fig:supp_megamontage_pc2_9}. Axes in UMAP and t-SNE are arbitrary. Since the algorithms prioritize local distances, long distances between clusters are not meaningful.
    ACB, African Caribbean in Barbados;
    ASW, African Ancestry in Southwest US;
    BEB, Bengali;
    CDX, Chinese Dai;
    CEU, Utah residents with Northern/Western European ancestry;
    CHB, Han Chinese;
    CHS, Southern Han Chinese;
    CLM, Colombian in Medellin, Colombia;
    ESN, Esan in Nigeria;
    FIN, Finnish;
    GBR, British in England and Scotland;
    GWD, Gambian;
    GTH, Gujarati;
    IBS, Iberian in Spain;
    ITU, Indian Telugu in the UK;
    JPT, Japanese;
    KHV, Kinh in Vietnam;
    LWK, Luhya in Kenya;
    MSL, Mende in Sierra Leone;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PJL, Punjabi in Lahore, Pakistan;
    PUR, Puerto Rican;
    STU, Sri Lankan Tamil in the UK;
    TSI, Tuscani in Italy;
    YRI, Yoruba in Nigeria}
    \label{fig:1000g_compare}
\end{figure*}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{0.49\columnwidth}
    \includegraphics[width=\columnwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.jpeg}
    \caption{}
    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
    \end{subfigure}\hfill
    \unskip\ \vrule\
        \begin{subfigure}{0.49\columnwidth}
    \includegraphics[width=\columnwidth]{images/Asthma_FullAsian_PC_QC_new_UMAP_PC5_NC2_NN15_MD05_201931173256_ethcob.jpeg}
    \caption{}
        \label{fig:umap_ukbb_indian}
    \end{subfigure}
    \caption{\textbf{Applying UMAP to subsets of data can reveal deep population structure.} \textbf{(a)} UMAP on the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster. Colouring the points by birthplace shows they were born almost entirely in the Mountain region (in green) of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana). When populations from the 1KGP are projected onto the UMAP embedding they do not map to the cluster. Six 1KGP populations are presented:
    CLM, Colombian in Medellin, Colombia;
    IBS, Iberian in Spain;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PUR, Puerto Rican;
    TSI, Tuscani in Italy. Figs~\ref{fig:supp_umap_hrs_hisp_admix} and \ref{fig:supp_umap_hrs_hisp_birth} present the same projection of individuals from the HRS coloured by estimated admixture proportions census region of birth, respectively. \textbf{(b)} UMAP on the top $8$ principal components of the self-identified Asian populations of the UKBB create clusters. Indian individuals born in Kenya (in purple) form one such cluster. A version coloured by self-identified ethnicity is presented in fig~\ref{fig:supp_umap_ukbb_asian_eth}. }
    \label{fig:subpops}
\end{figure*}

\subsection*{Results}
\subsubsection*{Fine-scale visualization of the 1KGP dataset} The 1KGP contains genotype data of 3,450 individuals from 26 relatively distinct labeled populations\cite{10002015global}. Fig~\ref{fig:1000g_compare} shows visualizations using PCA, t-SNE, UMAP, and UMAP with PCA pre-processing. Using UMAP and t-SNE on the genotype data presents clusters that are roughly grouped by continent, with UMAP showing a clear hierarchy of population and continental clusters, whereas t-SNE fails to assign many individuals to population clusters. Using either method on the top principal components leads to distinct population clusters and less defined continental structure. Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; adding further components converges to results similar to using the entire genotype data (see figs~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}). To investigate the population information contained in low-variance PCs, we performed UMAP on data projected onto PCs 100 to 3450 (i.e., without information about the leading 99 PCs). Fig~\ref{fig:supp_1kgp_3350} shows that population structure is still clearly visible.

Focusing on UMAP with the leading 15 principal components (fig~\ref{fig:1000g_compare}~(iv)), several population clusters reflect shared ancestries. British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry. Toscani and Iberian individuals form a cluster reflecting their Mediterranean heritage. African Americans in the Southwest US, African Caribbean individuals in Barbados, and some Puerto Ricans also form a cluster. Three East Asian clusters appear: one is largely Han and Southern Han individuals, another is comprised of the Chinese Dai in southern China and the Kinh from Vietnam, and the third is the Japanese population. Other clusters are comprised of Colombians and Peruvians, the Esan and Yoruba populations of Nigeria, and several South Asian populations.

Within population clusters, family members were projected near each other within broader population groups. When UMAP was parameterized to use only 5 nearest neighbours, however, families often formed distinct clusters (fig~\ref{fig:supp_1kgp_families}): Using few neighbors to build a manifold emphasizes closer relatedness.

A few individuals cluster with populations different from their label: some Mexican individuals cluster with Spanish and Italian populations; some Puerto Rican individuals cluster with African American and Caribbean populations; and one Gambian individual clusters with the Mende of Sierra Leone. Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering effect is robust to the number of components considered (fig~\ref{fig:supp_megamontage_pc10_50}). Differentiation in the 1KGP Gujarati population has been previously identified through a PCA restricted to the Gujarati \cite{reich2009india}. Following a preprint version of the present article, 23andme released a statement \cite{23andme} arguing that one of the two clusters could be traced, via 23andMe participant recontact to individuals from a group in Western India with shared ancestry and patronym. [D. Poznik, 23andme, personal communication, and \cite{23andme}].

@@

The HRS forms several large clusters, reflecting both ethnicity (fig~\ref{fig:supp_umap_hrs_eth}) and admixture proportions (fig~\ref{fig:umap_hrs_admix}). Gradients in admixture proportion are visible within the predominantly Hispanic cluster, but not within the predominantly Black cluster, perhaps because the variance in ancestry proportions is greater among Hispanics. The "White Not Hispanic" (WNH) group forms several interconnected clusters, and these do not correspond to broad geographical areas (fig~\ref{fig:supp_hrs_born}). By generating the PC axes and UMAP embedding for the HRS data in fig~\ref{fig:supp_umap_hrs_eth}, and projecting the 1KGP data onto it, we reveal substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population (fig~\ref{fig:supp_hrs_1kgp_projected}). One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.

\subsubsection*{Regional patterns in the Hispanic subpopulation} Applying UMAP to self-identified Hispanic individuals in the HRS reveals clear groupings related to birth region in fig~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}. The highlighted cluster consists almost entirely of individuals born in the Mountain Region of the United States. This cluster is not apparent when looking at a grid of pairwise plots of the first 8 principal components, provided in fig~\ref{fig:supp_hrs_hisp_grid}, as the signal is distributed along PCs 3, 4, and 6. Even though continental admixture patterns do correlate with UMAP position (fig~\ref{fig:supp_umap_hrs_hisp_admix}), these do not explain the Mountain Region cluster. Individuals from 1KGP populations do not appear in the cluster when projected to the UMAP embedding. The cluster possibly comprises the Hispano/Nuevomexicano population of the Southwest US, who have been present in the Mountain Region area long before the more recent immigrants from Latin America, and whose ancestry is expected to reflect both distinct Native ancestry and population-specific drift relative to other Hispanic populations. Such a cluster has been previously identified in AncestryDNA data using network-based clustering on identity-by-descent connections\cite{han2017clustering}; a recent preprint discusses the Mountain Region Hispanics with a more detailed historical description\cite{Jordan333609}.

\subsubsection*{Population structure in the UKBB reflects local and global genetic variation} 
The UKBB contains data on 488,377 individuals including genotypes, phenotypic measures and self-identified ethnic backgrounds. Fig~\ref{fig:fig_ukbb} compares UMAP to PCA applied to UKBB. As expected, PCA captures major axes of variation emphasizing continental ancestry, whereas UMAP reveals finer structure. UMAP on the top 10 principal components reveals continuous and discrete population structure (fig~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies multiple sub-populations, as well as continuous structure within populations and admixture gradients between populations. %The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of a large and multi-ethnic dataset.
The result is a succinct illustration of the complex structure and population relationships in a large and multi-ethnic dataset.

@@

The largest cluster in fig~\ref{fig:umap_ukbb_pc10} consists of the White British and Irish populations. The Irish population forms a sub-cluster, but many individuals are also scattered throughout the British-identifying population. Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters. Chinese individuals form a cluster, within a broader East Asian population; Indian, Pakistani, and Bangladeshi populations form a closely bound cluster as well. The East Asian and South Asian populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group". The patchwork of genetic neighbourhoods is connected by trails of admixed individuals, which converge at a nexus of individuals with a variety of ethnicities. Many claim mixed ancestry, and there are clusters of individuals who belong to an "other ethnic group". Using data on countries of birth, we identified many finer groups in fig~\ref{fig:supp_ukbb_cob}, and confirmed they appeared in intuitive areas with, e.g., Japanese and Filipino clusters being projected near Chinese clusters.

Fig~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection from fig~\ref{fig:umap_ukbb_pc10} coloured instead by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly. Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe and British-wide data\cite{novembre2008europe,leslie2015fine}. Fig~\ref{fig:umap_ukbb_pc10} shows that the admixed individuals have UMAP coordinates next to White British individuals residing in the South East of the UK, where London is located (see also fig~\ref{fig:supp_london_distance}, where individuals are coloured by distance from London). This likely reflects the high migration levels to the city and surrounding area: the UMAP projection attempted to preserve both the genetic similarity among admixed individuals and the relatedness with White British individuals in cosmopolitan areas.

 The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved. Figs~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB.

\begin{figure}
    \centering
    \begin{subfigure}{\columnwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ns_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ew_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \caption{}
    \label{fig:umap_ukbb_geo_osgb}
    \end{subfigure}
     \begin{subfigure}{\columnwidth}
         \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_f.pdf}
    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
   \caption{}
    \label{fig:umap_height_female}
      \end{subfigure}
      \caption{\textbf{UMAP captures relationships between population structure, geography, and phenotype heterogeneity.} \textbf{(a)} Each individual is coloured by their geographical coordinates of residence. Coordinates follow the UKBB's OSGB1936 geographic grid system and represent distance from the Isles of Scilly, which lie southwest of Great Britain. The left image colours individuals by their north-south ("northing") coordinates, and the right image colours them by their east-west ("easting") coordinates. Adding more components creates finer clusters.  (figs~\ref{fig:supp_montage_ukbb_ns} and~\ref{fig:supp_montage_ukbb_ew}). Northing values were truncated between $100$km and $700$km, and easting values were truncated between $200$km and $600$km. \textbf{(b)} Females from the same projection coloured by age-adjusted difference from mean population height (left) and leukocyte counts (right). Individuals with missing data were excluded. To protect participant privacy, data in these images has been randomized as explained in the materials and methods section.}
\end{figure}

As an alternate visualization of geography and genetic diversity, we performed a 3D UMAP projection and converted the UMAP coordinates into RGB values, allowing us to plot individuals on a map of Great Britain, emphasizing both spatial gradients of genetic relatedness and increased diversity in urban centers (fig~\ref{fig:umap_ukbb_geo_map}). The geographical patterns outside major urban centers are similar to those reported in~\cite{leslie2015fine} using the haplotype-based CHROMOPAINTER on British individuals whose grandparents lived nearby. Using data about country of birth, we performed a similar analysis of a world map~\ref{fig:umap_world_map}, revealing subtle regional variation around the world.
